---
title: "Analyzing Financial and Economic Data with R"
subtitle: "Chapter 06 - Dataframes and other objects"
author: "Marcelo S. Perlin"
date: "2020-02-15"
output:
  slidy_presentation: default
  ioslides_presentation: default
  beamer_presentation: default

---

## Introduction {#introduction}

```{r, include=FALSE}
my.fig.height <- 3
my.fig.width <- 4
my.out.width <- '100%'
book.site.zip <- 'https://www.msperlin.com/blog/static/afedr-files/afedr_files.zip'

format.cash <- function(x) {
  require(scales)

  x.formatted <- dollar(x,
                        prefix = '$',
                        decimal.mark = '.',
                        big.mark = ',',
                        largest_with_cents = Inf)

  return(x.formatted)
}

```

- In R, everything is an object with its own properties

- Lists, Dataframes and matrices help organize other simpler classes of objects


## `Dataframes`

A `dataframe` is simply a table with rows and columns, and integrates nicelly with the `tidyverse`. 

**A `dataframe` can organize our work significantly**. The tabular structure of a `dataframe` forces the data to be _paired_, where each row is a different data point with several pieces of information (columns).


### Creating `dataframes`

We call function `tibble::tibble` to create a `dataframe`:


```{r, tidy=FALSE}
library(tidyverse)

# set tickers
tickers <- c(rep('AAP',5), 
             rep('COG', 5), 
             rep('BLK', 5), 
             rep('CAM',5) )

# set a date vector
dates <- as.Date(rep(c("2010-01-04", "2010-01-05", "2010-01-06", 
                       "2010-01-07", "2010-01-08"), 4) )

# set prices					  
prices <- c(40.38,  40.14,  40.49,  40.48,  40.64,
            46.23,  46.17,  45.97,  45.56,  45.46,
            238.58, 239.61, 234.67, 237.25, 238.92,
            43.43,  43.96,  44.26,  44.5,   44.86)

# create tibble/dataframe
my_df <- tibble(tickers, dates, prices)

# print its first 5 rows
print(head(my_df))
```


### Inspecting a Dataframe

Once you have a dataframe in your R session, the following and **mandatory** step is to check its contents. We need to be aware of these items:

- Properly defined column's names and classes;
- Correct number of rows and columns;
- The existence (or not) of missing data (`NA`).


One of the most recommended functions for familiarizing yourself with a `dataframe` is `dplyr::glimpse`.

```{r}
# check content of my_df
glimpse(my_df)
```

Usually, the use of `glimpse` is sufficient to check if the data import process has succeeded. However, a deeper analysis can also help to understand how each column varies in the imported data and if that makes sense to our problem. Here comes the role of the `base::summary` function:

```{r}
# check variation my_df
summary(my_df)
```


### The _pipeline_ Operator (`%>%`)

The _pipeline_ operator allows data operations to be performed sequentially and in a modular fashion, increasing readability and maintainability of the resulting code. 

Imagine a situation where we have three functions to be applied to a `dataframe`. Each function depends on the output of another function. This requires chaining your calls, so one output feeds an object to the next call. Using the _pipeline_ operator, we can write the `dataframe` manipulation procedure with the following code: \index{dplyr!\%>\%}

```{r, eval=FALSE}
my_tab <- my_df %>%
  fct1(arg1) %>%
  fct2(arg2) %>%
  fct3(arg3)
```

To compare, here are two ways to perform the same operation **without** the _pipeline_ operator:

```{r, eval=FALSE}
# version 1
my_tab <- fct3(fct2(fct1(my_df,
                         arg1),
                    arg2),
               arg1)

# version 2
temp1 <- fct1(my_df, arg1)
temp2 <- fct2(temp1, arg2)
my_tab <- fct3(temp1, arg3)
```


### Accessing Columns

To discover the names of the columns of a `dataframe`, we have two functions, `names` and `colnames`, with the exact same behavior:  \index{base!names} \index{base!colnames}

```{r}
# get names of columns with names
names(my_df)
colnames(my_df)
```

Both can also modify column names:

```{r}
# set temp df
temp_df <- my_df

# change names
names(temp_df) <- paste('Col', 1:ncol(temp_df))

# check names
names(temp_df)
```

To access a particular column of a `dataframe`, we can use operator `$` or the name/position of the column with brackets. See next:

```{r}
# isolate columns of df
my_tickers <- my_df$tickers
my_prices <- my_df['prices']
my_dates <- my_df[ ,2]

# print the results
print(head(my_tickers))
print(head(my_prices))
print(head(my_dates))
```

To access specific rows and columns of a `dataframe`, use single brackets with atomic vectors that indicate positions:

```{r}
# accessing rows 1:5, column 2
print(my_df[1:5, 2])

# accessing rows 1:5, columns 1 and 2
print(my_df[1:5, c(1,2)])

# accessing rows 1:5, all columns
print(my_df[1:5, ])
```

Column selection can also be performed using names, as in the following example:

```{r}
# selecting rows 1 to 3, columns 'ticker' and 'prices'
print(my_df[1:3, c('tickers', 'prices')])
```

Or, using the pipeline operator and function  `dplyr::slice` and `dplyr::select`: \index{dplyr!select}

```{r}
my.temp <- my_df %>%
  select(tickers, prices) %>%
  slice(1:3) %>%
  glimpse()
```


### Modifying a `dataframe`

To create new columns in a dataframe, simply use function `dplyr::mutate` and the _pipeline_ operator. \index{adding column to dataframe} \index{dplyr!mutate}

```{r}
# add columns with mutate
my_df <- my_df %>%
  mutate(ret = prices/lag(prices) -1,
         seq_1 = 1:nrow(my_df),
         seq_2 =  seq_1 +9) %>%
  glimpse()
```

Another, more traditional way of creating new columns is using operator `$`:

```{r}
# add new column with base R
my_df$seq_3 <- 1:nrow(my_df)

# check it
glimpse(my_df)
```

Therefore, you can use `$` to either access or modify a dataframe. 

Going further, if we try to create a column with the number of elements different than the number of rows of the target `dataframe`, an error will appear. 

```{r,error=TRUE}
my_df <- my_df %>%
  mutate(seq_3 =  1:100) %>%
  glimpse()
```

However, due to the simplified recycling rule, we can use single values to fill up a whole column: 

```{r}
my_df <- my_df %>%
  mutate(seq_3 =  1) %>%
  glimpse()
```

To remove columns from a `dataframe`, use function `dplyr::select` with the minus symbol for the undesired columns:

```{r}
# removing columns
my_df.temp <- my_df %>%
  select(-seq_1, -seq_2, -seq_3) %>%
  glimpse()
```

Using base R, the traditional way of removing columns is to allocate a single value `NULL` to its contents:

```{r}
# set temp df
temp_df <- my_df

# remove cols
temp_df$prices <- NULL
temp_df$dates  <- NULL
temp_df$ret  <- NULL
temp_df$tickers  <- NULL

# check it
glimpse(temp_df)
```


### Filtering rows of a `dataframe`

A fairly common `dataframe` operation in R is to filter rows according to one or more conditions. For example, if we only wanted data from the `'COG'` stock, we could use the `filter` function to filter the table: \index{dplyr!filter}

```{r}
# filter df for single stock
my_df.temp <- my_df %>%
  filter(tickers == 'COG') %>%
  glimpse()
```

We can go further and also filter data for `'COG'` and dates after `'2010-01-05'`:

```{r}
# filter df for single stock and date
my_df.temp <- my_df %>%
  filter(tickers == 'COG',
         dates > as.Date('2010-01-05')) %>%
  glimpse()
```

Here we used symbol `==` to test for equality in column `ticker` and _greater than_ (>) for selecting the rows where the dates are after 2010-01-05. 


### Sorting a `dataframe`

As an example, consider creating a `dataframe` with these values:

```{r, tidy=FALSE}
# set new df
my_df <- tibble(col1 = c(4, 1, 2), 
                col2 = c(1, 1, 3), 
                col3 = c('a','b','c'))

# print it					
print(my_df)
```

We use function  `dplyr::arrange` and the _pipeline_ operator to order the whole `dataframe` by the ascending values of column `col1`:

```{r}
# sort ascending, by col1 
my_df <- my_df %>%
  arrange(col1) %>%
  print()
```

We can also sort by descending values using `desc`:

```{r}
# sort descending, col1 and col2
my_df <- my_df %>%
  arrange(desc(col1)) %>%
  print()
```

And, for multiple columns, using extra arguments in `arrange`: 

```{r}
# sort ascending, by col2 and col1
my_df <- my_df %>%
  arrange(col2, col1) %>%
  print()
```

As for base R, function `order` returns the position of the elements for the sorted vector. With the first column of `my_df`, the positions of the elements in ascending order are:

```{r}
# set index with positions of ascending order in col1
idx <- order(my_df$col1)

# print it
print(idx)
```


### Combining and Aggregating `dataframes`

In the practice of manipulating data, often you must aggregate multiple `dataframes` into a single one. For that, we have functions `dplyr::bind_rows` (alternative to `base::rbind`) and `dplyr::bind_cols` (alternative to `base::cbind`). Examples of usage are given next. \index{dplyr!bind\_rows} \index{dplyr!bind\_cols}

```{r}
# set two dfs with same colnames
my_df_1 <- tibble(col1 = 1:5, 
                  col2 = rep('a', 5))
my_df_2 <- tibble(col1 = 6:10, 
                  col2 = rep('b', 5))

# bind them by rows
my_df <- bind_rows(my_df_1, my_df_2)

# print result
print(my_df)
```


Another interesting aspect of `dplyr::bind_rows` is that, if the names of the columns don't match, the unmatched columns will return a `NA` result (not available). 

```{r, eval=TRUE, tidy=FALSE}
# set two df with different colnames
my_df_1 <- tibble(col1 = 1:5, 
                  col2 = rep('a', 5))
my_df_2 <- tibble(col1 = 6:10, 
                  col3 = rep('b', 5))

# bind them by rows (NA values for missing cols)
my_df <- bind_rows(my_df_1, 
                   my_df_2)

# print result
print(my_df)
```

For the case of column bind with function `dplyr::bind_cols`, the names of the columns must be different, but the number of rows must be the same: \index{dplyr!bind\_cols}

```{r}
# set two dfs
my_df_1 <- tibble(col1 = 1:5, 
                  col2 = rep('a', 5))
my_df_2 <- tibble(col3 = 6:10, 
                  col4 = rep('b', 5))

# column bind dfs
my_df <- cbind(my_df_1, my_df_2)

# print result
print(my_df)
```

Sometimes, aggregating different tables won't be as easy as simply row or column binding. 

For that, you can use functions `dplyr::join*` to merge two dataframes using one or more indexes. 

For example, `dplyr::inner_join` keeps only the data where it finds a matching index, while `dplyr::full_join` keep everything from both tables and fills the missing information with `NA` values. Let's have a closer look by using a practical example. 

```{r}
# set df
my_df_1 <- tibble(date = as.Date('2016-01-01')+0:10,
                  x = 1:11)

my_df_2 <- tibble(date = as.Date('2016-01-05')+0:10,
                  y = seq(20,30, length.out = 11))

```

Please do notice that both dataframes share a column called `date`, which will be automatically used as a matching index. 

```{r}
# aggregate tables
my_df <- inner_join(my_df_1, 
                    my_df_2)

glimpse(my_df)
```

Now with `dplyr::full_join`:

```{r}
# aggregate tables
my_df <- full_join(my_df_1, 
                   my_df_2)

glimpse(my_df)
```

Notice the difference in the number of rows from one to the other. When using `dplyr::full_join`, all unmatched cases are set to `NA`.


## `Lists`

A `list` is a flexible container that can hold many elements. Unlike atomic vectors, a `list` has no restriction on the classes or types of elements.


### Creating `lists`

A `list` can be created with the `base::list` command, followed by their comma-separated elements: \index{base!list}

```{r}
# create list
my_l <- list(c(1, 2, 3),
             c('a', 'b'),
             factor('A', 'B', 'C'),
             data.frame(col1 = 1:5))

# use base::print
print(my_l)

# use dplyr::glimpse
glimpse(my_l)
```

Following other objects, the elements of a `list` can also be named:

```{r}
# set named list
my_named_l <- list(tickers = 'CMPY',
                   markets = 'NYSE',
                   df_prices = data.frame(P = c(1,1.5,2,2.3),
                                          ref_date = Sys.Date()+0:3))

# check content
glimpse(my_named_l)
```

The information is self-contained in a single object. By using a `list` object, we can aggregate and structure all sorts of information about the company `CMPY`. 


### Accessing the Elements of a `list`

As mentioned, the individual elements of a `list` can be accessed with double brackets ( `[[ ]]`), as in: \index{[[ ]]}

```{r}
# accessing elements from list
print(my_named_l[[2]])
print(my_named_l[[3]])
```

You can also access the elements of a `list` with simple brackets (`[]`):

```{r}
# set list
my_l <- list('a',
             c(1, 2, 3),
             factor('a', 'b'))

# check classes
class(my_l[[2]])
class(my_l[2])
```

If we try to add an element to `my_l[2]`, we will receive an error message.

```{r error=TRUE}
# adding an element to a list (WRONG)
my_l[2] + 1
```

An error is returned because a `list` object cannot be summed with a `numeric` object. To fix it, simply use double brackets, as in `my_l[[2]] + 1`. 


### Adding and Removing Elements from a `list`

To add or replace elements in a `list`, just set the new object in the desired position:

```{r}
# set list
my_l <- list('a', 1, 3)
glimpse(my_l)

# add new elements to list
my_l[[4]] <- c(1:5)
my_l[[2]] <- c('b')

# print result
glimpse(my_l)
```

This operation is also possible with the use of names and `$`:

```{r}
# set list
my_l <- list(elem1 = 'a', 
             name1=5)

# set new element
my_l$name2 <- 10

# check it
glimpse(my_l)
```

To remove elements from a `list`, set the element to the reserved symbol `NULL`, as in: \index{NULL}

```{r}
# set list
my_l <- list(text = 'b', num1 = 2, num2 = 4)
glimpse(my_l)

# remove elements
my_l[[3]] <- NULL
glimpse(my_l)

# remove elements
my_l$num1 <- NULL
glimpse(my_l)
```

Another way of removing elements from a `list` is to use a negative index, which will exclude it from the returned object. 

```{r}
# set list
my_l <- list(a = 1, 
             b = 'text')

# remove second element
glimpse(my_l[[-2]])
```

As with atomic vectors, removing elements of a `list` can also be accomplished with logical conditions. See next:

```{r}
# set list
my_l <- list(1, 2, 3, 4)

# remove elements by condition
my_l[my_l > 2] <- NULL
glimpse(my_l)
```

However, note this operation only works because all elements of `my_l` are numeric, and a logical test can be applied to all cases. If that is impossible for a particular element, R will return an `NA` value. 



## `Matrices`

A matrix is a two-dimensional representation of numbers, arranged in rows and columns. Using matrices is a powerful way of representing numerical data in two dimensions.

A simple example of using matrices in finance is the representation of stock prices over time. The rows of the matrix represents the different dates, and the columns set each stock apart:

```{r, echo=FALSE, results='asis', purl=FALSE}
my_f <- afedR::afedR_get_data_file('SP500-Stocks_wide.csv')
temp_df  <- read_csv(my_f, 
                     col_types = cols())

temp_df$Date <- temp_df$date 
rownames(temp_df) <- temp_df$Date

selected_assets <- c('AAP', 'COG', 'BLK', 'CAM')

temp_df <- temp_df[1:5, selected_assets]
knitr::kable(temp_df, format = 'markdown', digits = 2, 
             row.names = T )
```

The above matrix could be created in R with the following code:

```{r, tidy=FALSE}
# set raw data with prices 
raw_data <- c(40.38,  40.14,  40.49,  40.48,  40.64,
              46.23,  46.17,  45.97,  45.56,  45.46,
              238.58, 239.61, 234.67, 237.25, 238.92,
              43.43,  43.96,  44.26,  44.5,   44.86)

# create matrix          
my_mat <- matrix(raw_data, nrow = 5, ncol = 4)
colnames(my_mat) <- c('AAP', 'COG', 'BLK', 'CAM')
rownames(my_mat) <- c("2010-01-04", "2010-01-05", "2010-01-06", 
                      "2010-01-07", "2010-01-08")

# print result
print(my_mat)
```

We set the number of rows and columns explicitly with arguments `nrow = 4` and `ncol = 3` in `base::matrix`. The names of rows and columns are defined with functions `colnames` and `rownames`, using a left side notation as in `rownames(my_mat) <- c(...)`. 

Going further, we can also retrieve the names of rows and columns with the same functions: \index{base!rownames} \index{base!colnames} 

```{r}
# print the names of columns 
print(colnames(my_mat))

# print the names of rows
print(rownames(my_mat))
```

A `matrix` object is also flexible with its content, as long as it is a single class. For example, you can create matrices with `character` elements:

```{r, tidy=FALSE}
# create matrix with character
my_mat_char <- matrix(rep(c('a','b','c'), 3), 
                      nrow = 3, 
                      ncol = 3)

# print it					  
print(my_mat_char)
```

Now with a `logic` type:

```{r, tidy=FALSE}
# create matrix with logical
my_mat_logical <- matrix(sample(c(TRUE,FALSE), 
                                size = 3*3,
                                replace = TRUE),
                         nrow = 3, 
                         ncol = 3)

# print it					  
print(my_mat_logical)
```						 

This flexibility allows the user to expand the representation of two-dimensional data beyond numerical values.


### Selecting Elements from a `matrix`

Following the same notation as the atomic vector, you can select _pieces_ of a `matrix` using indexes.

```{r}
# create matrix
my_mat <- matrix(1:9, nrow = 3)

# display it
print(my_mat)

# display element in [1,2]
print(my_mat[1,2])
```

To select an entire row or column, simply leave a blank index, as in the following example:

```{r}
# select all rows from column 2
print(my_mat[, 2])

# select all columns from row 1
print(my_mat[1, ])
```

Notice the result of indexing is an atomic vector, not a `matrix`. If we wanted the extracted piece to maintain its `matrix` class, with vertical or horizontal orientation, we could force this conversion using functions `as.matrix` and `matrix`: \index{base!matrix}  \index{base!as.matrix} 

```{r}
# force matrix conversion and print result
print(as.matrix(my_mat[ ,2]))

# force matrix conversion for one row and print result
print(matrix(my_mat[1, ], nrow=1))
```

Pieces of the `matrix` can also be selected using vectors. If we wanted a new `matrix` with all elements from the second row and first column to the third row and second column, we could use the following code:

```{r}
# select some elements and print them
print(my_mat[2:3, 1:2])
```

Finally, using logical tests to select elements of matrices is also possible: \index{logical matrices} 

```{r}
# set matrix
my_mat <- matrix(1:9, nrow = 3)

# print logical matrix where value is higher than 5
print(my_mat >5)

# print the result
print(my_mat[my_mat >5])
```


## Exercises 

01. Use functions from the `tibble` package to create a dataframe called `my_df` with a column named `x` containing a randomized selection (see function `sample`) of a sequence from 1 to 100 and another column named `y` with the value of column x plus 5. How many values in column `y` are greater than 10 and less than 25?

02. Using operator `$`, create a new column named `cumsum_x` in `my_df`, which will contain the cumulative sum of `x` (see `cumsum` function). How many values of the new column `cumsum_x` are greater than 50?

03. Use function `dplyr::filter` and the _pipeline_ operator to filter `my_df`, so it only keeps rows where the value of column `x` is greater than 30.

04. If not done yet, repeat exercises 1, 2 and 3 using the `tidyverse` functions and the _pipeline_ operator to chain all operations.

05. Use package `BatchGetSymbols` to download Facebook (FB) stock data since `'2010-01-01'` until today. If an investor bought `r format.cash(1000)` worth of FB stock on the first available day after `'2010-01-01'`, what would be the value of his portfolio today?

06. Use function `afedR::afedR_get_data_file` to find the path of the book file `'grunfeld.csv'`. After reading it with `readr::read_csv`, use functions `dplyr::glimpse` and `base::summary` to better understand the content of the imported dataset.

07. Create a list object with three dataframes in its contents, `df1`,` df2` and `df3`. The content and size of dataframes is your choice. Use `sapply` function to discover the number of rows and columns in each dataframe.

08. Create an identity matrix (value 1 diagonally, zero anywhere else) of size 3 X 3. Tip: Use function  `diag` to set the diagonal values in the matrix. 
