---
title: "Analyzing Financial and Economic Data with R"
author: "Marcelo S. Perlin"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
  word_document: default
subtitle: "Chapter 05 - Importing Data from the Internet"
---

```{r setup, include=FALSE}
book.site <- 'https://www.msperlin.com/blog/publication/2017_book-pafdr-en/'
book.site.zip <- 'https://www.msperlin.com/blog/files/pafdr%20files/Code_Data_pafdR.zip'

knitr::opts_chunk$set(message = FALSE)
```


## Exercises

**01. Using the `BatchGetSymbols` package, download price data for Tesla, Inc. (TSLA) in the last two years. What is the lowest adjusted price you can find? What is the date with the lowest price? Tip: Use function `which.min` to find the position of the minimum price. **

```{r}
library(BatchGetSymbols)

ticker <- 'TSLA'
first_date <- Sys.Date() - 2*365
last_date  <- Sys.Date()

df_prices <- BatchGetSymbols(tickers = ticker, 
                             first.date = first_date, 
                             last.date = last_date)[[2]]

my_msg <- paste0('The lowest price is ', min(df_prices$price.adjusted), ' at ', 
                 df_prices$ref.date[which.min(df_prices$price.adjusted)], '.')
cat(my_msg)
```


**2. Use `tidyquant::tq_get` to download historical values for the SP500 index (^GSPC) since 1950 until today. What is the annual compound rate of return for the index?**


```{r}
library(tidyquant)

ticker = '^GSPC'
df_prices <- tq_get(x = ticker, get = 'stock.prices',
                    from = '1950-01-01',
                    to = Sys.Date())


n_years <- (max(df_prices$date) - min(df_prices$date))[[1]]/365
total_ret <- df_prices$adjusted[nrow(df_prices)]/df_prices$adjusted[1] - 1 

annual_rate <- (1+total_ret)^(1/n_years) - 1

my_msg <- paste0('The annual compound rate of ', ticker, ' (',
                 df_prices$date[1], '::', df_prices$date[nrow(df_prices)],
                 ') is ', scales::percent(annual_rate), ' per year.')
cat(my_msg)
```

**03. Use the [Quandl search box](https://www.quandl.com/search)^[https://www.quandl.com/search] to find available databases for your country. This includes databases from Central Banks and research agencies. Cite those that are most popular and most likely to be used in your work. **

In Quandl website I find two important databases for my country: Central Bank of Brazil Statistical Database and CEPEA (Center for Applied Studies on Applied Economics (Brazil)). 

**04. Create a profile on the [Quandl website](https://www.quandl.com/)^[https://www.quandl.com/]. Using `GetQuandlData`, download unemployment data for Brazil using the Central Bank of Brazil Statistical Database. What is the latest value of this series? **

```{r}
library(GetQuandlData)

id_series <- 'BCB/24369'
my_api_key <- readLines('~/Dropbox/.quandl_api.txt')

df_unemp_br <- get_Quandl_series(id_in = id_series, 
                                 api_key = id_series)

my_msg <- paste0('The latest value for ', id_series, ' is ', 
                 df_unemp_br$value[which.max(df_unemp_br$ref_date)], ' at ',
                 df_unemp_br$ref_date[which.max(df_unemp_br$ref_date)], '.')
cat(my_msg)
```


**04. What is the latest value of EUR Bitcoin at BTC? Use trade price data information from the `Kraken` exchange to solve the exercise.**

```{r}
library(Rbitcoin)

# set mkt, currency pair and type of action
mkt <- "kraken"
currency <- c("BTC","EUR")
action <- 'trades'

Quandl.api_key('Esv7Ac7zuZzJSCGxynyF')
# import data
my_l <- market.api.process(market = mkt,
                           currency_pair = currency,
                           action = action)

# print it
trades_btc <- my_l$trades

my_msg <- paste0('The latest price of bitcoin in ', currency[1], '/', currency[2], 
                 ' is ', trades_btc$price[which.max(trades_btc$date)], '.')
cat(my_msg)
```
