---
title: "Analyzing Financial and Economic Data with R"
author: "Marcelo S. Perlin"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
  word_document: default
subtitle: "Chapter 07 - Basic Object Classes"
---

```{r setup, include=FALSE}
book.site <- 'https://www.msperlin.com/blog/publication/2017_book-pafdr-en/'
book.site.zip <- 'https://www.msperlin.com/blog/files/pafdr%20files/Code_Data_pafdR.zip'

knitr::opts_chunk$set(message = FALSE)
```


## Exercises

**01. Let's assume that, on a certain date, you bought 100 shares of a company, 15 dollars per share. After some time, you sold 30 shares for 18 dollars each and the remaining shares were sold for 22 dollars in a later day. Using a _script_ in R, structure this financial problem by creating numeric objects. What is the gross profit of this stock market transaction? **

```{r}
total_shares <- 100
price_purchase <- 15

total_purchase_value <- total_shares*price_purchase

total_sell_1 <- 30*18
total_sell_2 <- (total_shares-30)*22

total_sell_value <- total_sell_1 + total_sell_2

# solution
gross_profit <- total_sell_value - total_purchase_value

print(gross_profit)
```


**02. Consider these numeric vectors `x` and `y`:**

```{r}
set.seed(7)
x <- sample (1:3, size = 5, replace = T)
y <- sample (1:3, size = 5, replace = T)
```

**What is the sum of the elements of a new vector resulting from the multiplication between the elements of x and y?**

```{r}
sum(x*y)
```


**03. Create a sequence called `seq_1`, with values between -10 and 10, where the interval between elements is equal to 2.**

```{r}
seq_1 <- seq(from = -10, to = 10, by = 2)

seq_1
```


**04. Define another object named `seq_2`, which contains a sequence of length 1000, with values between 0 and 100.**

```{r}
seq_2 <- seq(from = 0, 
             to = 100, 
             length.out = 1000)

seq_2
```

**05. Calculate the sum between `seq_1` and `seq_2`. Did this operation work despite the different sizes of the vectors? Explain your answer.**

```{r}
library(tidyverse)

test <- seq_1 + seq_2

glimpse(test)
```

Yes, it worked, but with a warning: _"longer object length is not a multiple of shorter object length"_. The explanations is that, whenever R find operations with vector of different sizes, it uses the recycling rule, where the shortest vector is repeated the number of times as needed in order to match the size of the longer vector.

**06. If we create an object with the cumulative sum of a sequence from 1 to 100, what is the position of the first element that exceeds the value of 50?**

```{r}
my_sum <- cumsum(1:100)

# solution
print(which(my_sum > 50)[1])
```


**07. Create a vector according to the following formula, where $i=1 ... 100$. What is the value of the sum of the elements of x? **

$$
x_i = \frac{(-1^{i + 1})}{2i-1}
$$


```{r}
i <- 1:100

x <- (-1^(i+1))/(2*i - 1)

glimpse(x)
```


**08. CHALLENGE - Create a $z_i$ vector according to the following formula, where $x_i = 1...50$ and $y_i = 50...1$. What is the value of the sum of $z_i$ elements? (TIP: see function `dplyr::lag`)**

$$
z_i = \frac{y_i-x_{i-1}}{y_{i-2}}
$$

```{r}
x <- 1:50
y <- 50:1

# solution using `base`
z <- (y - c(NA, x[1:(length(x)-1)]))/c(NA, NA, y[1:(length(y)-2)])

# solution with tidyverse (much prettier huh!)
z <- (y - lag(x, n = 1))/lag(y, n = 2)

# solution (be aware of the NA values)
sum(z, na.rm = TRUE)
```


**09. Create an object named `x` with 1000 random values taken from a Normal distribution with average and standard deviation equal to 10. Using the `cut` function, create another object that defines two groups based on values of `x,` one for values higher than 10, and another for values under -5. How many observations do we find in each group? **

```{r}
x <- rnorm(n = 1000, mean = 10, sd = 10)

my_group <- cut(x, 
                breaks = c(-Inf, -5, 10, Inf))

# solution
table(my_group)
```


**10. Execute the following code and create object `my_char`: **

```{r}
set.seed(10)
my_char <- paste(sample(letters, 500, replace = T), 
                 collapse = '')
```


**How many times is the letter `'x'` found in the resulting text object?**

```{r}
library(tidyverse)

# solution
my_count <- str_count(my_char, 'x')
my_count 
```


**11. If we split the `my_char` object from previous exercise it into several smaller pieces of text using the letter `"b"`, what is the number of characters in the largest piece?**

```{r}
my_split <- str_split(my_char, pattern = 'b')[[1]]

# find number of characters in each
n_chars <- sapply(my_split, nchar)

# solution
n_chars[which.max(n_chars)]
```


**12. At link [https://www.gutenberg.org/ebooks/2264.txt.utf-8](https://www.gutenberg.org/ebooks/2264.txt.utf-8), you can find a file containing the full text of William Shakespeare's "Macbeth" book. Use the link directly with function `readr::read_lines` to import the entire book as a vector of characters called `my_book.` How many lines does the resulting object have? **

```{r}
my_link <- 'https://www.gutenberg.org/ebooks/2264.txt.utf-8'

my_book <- read_lines(my_link)

# solution
print(length(my_book))
```


**13. Aggregate the vector of characters in `my_book` to a single text object named `full_text` using function `paste0(my_book, collapse = '\n')`. With functions from package `stringr,` check how often the word `'lord'` repeats itself throughout the text. **

```{r}
full_text <- paste0(my_book, collapse = '\n')

# solution
my_count <- str_count(full_text, 'Bennet')
print(my_count)
```


**14. CHALLENGE - Use function `stringr::str_split` in object `full_text` to break the whole text as a function of whitespace (`' '`). Use this information with the function `base::table` and create a frequency table, showing the number of occurrences of each word. What are the five most used words in the text?**

```{r}
my_split <- str_split(full_text, pattern = fixed(' '))[[1]]

my_tab <- sort(table(my_split))

# solution
tail(my_tab, n = 5)
```

 
**15. Assuming you'll live for 100 years, what is the proportion of your birthdays that will land on a weekend day (Sunday/Saturday)? Tip: Use operator `%in%` to check a multiple equality condition.**

```{r}
b_day <- as.Date('1983-07-11')

b_day_vec <- seq(b_day, b_day + 100*365, by = '1 year')

w_days <- weekdays(b_day_vec)

my_prop <- scales::percent(sum(w_days %in% c('Saturday', 'Sunday'))/length(b_day_vec))
my_msg <- paste0('\nProportion of birthdays that lands on a weekend is ', 
                 my_prop, '.')
cat(my_msg)
```

 
