---
title: "Analyzing Financial and Economic Data with R"
author: "Marcelo S. Perlin"
date: "`r Sys.Date()`"
output:
  html_document: default
  word_document: default
  pdf_document: default
subtitle: 'Chapter 10 - Creating and Saving Figures with ggplot2'
---

```{r setup, include=FALSE}
book.site <- 'https://www.msperlin.com/blog/publication/2017_book-pafdr-en/'
book.site.zip <- 'https://www.msperlin.com/blog/files/pafdr%20files/Code_Data_pafdR.zip'

knitr::opts_chunk$set(message = FALSE)
```


## Exercises

**01. Download Facebook (FB) stock data  with the `BatchGetSymbols` package for the past 500 days. Display a line graph of the adjusted prices over time using the `ggplot2::ggplot` function. Make sure that:**

**- The x and y axis are correctly named;**
**- The plot has a title ("Prices for 1 stock"), subtitle ("Data from YYYY-MM-DD to YYYY-MM-DD") and a caption ("Beautiful solution of exercise 01, chapter 10.")**

```{r}
library(BatchGetSymbols)
library(tidyverse)

tickers <- 'FB'

df_FB <- BatchGetSymbols(tickers = tickers, 
                         first.date = Sys.Date() - 500,
                         last.date = Sys.Date())[[2]]

p <- ggplot(df_FB, aes(x = ref.date, y = price.adjusted)) + 
  geom_line() + 
  labs(title = paste0('Prices for ', length(tickers), ' stock') ,
       subtitle = paste0('Data from ', min(df_FB$ref.date), ' to ', 
                         max(df_FB$ref.date), '\n',
                         'Beautiful solution of exercise 01, chapter 10'),
       caption = 'Data from Yahoo Finance')

print(p)
```


**02. Download Google (GOOG), Facebook (FB) and Dropbox (DBX) stock data with `BatchGetSymbols` for the last 1500 days. Use function `ggplot` to create a single chart for all stock prices, where each ticker has a different line color. Maintain all other aspects of the plot from the previous exercise. **

```{r}
library(BatchGetSymbols)

tickers <- c('FB', 'DBX', 'GOOG') 

df_stocks <- BatchGetSymbols(tickers = tickers, 
                             first.date = Sys.Date() - 500,
                             last.date = Sys.Date())[[2]]

p <- ggplot(df_stocks, aes(x = ref.date, 
                           y = price.adjusted,
                           color = ticker)) + 
  geom_line() + 
  labs(title = paste0('Prices for ', length(tickers), ' stocks'),
       subtitle = paste0('Data from ', min(df_stocks$ref.date), ' to ', 
                         max(df_stocks$ref.date)),
       caption = 'Beautiful solution of exercise 01, chapter 10.')

print(p)
```

**03. For the previous chart, add points to the lines.**

```{r}
p <- p + geom_point()

print(p)
```


**04. For the same chart, separate stock prices into different panels with the `ggplot::facet_wrap` function. Also "release" the _y_ axis for the different stocks so that each stock has its own vertical axis.**

```{r}
p <- p + 
  facet_wrap(~ticker, scales = 'free')

print(p)
```


**05. Modify the previous chart theme to greyscale for both the chart area and the line colors.**

```{r}
p <- p + 
  theme_bw() + 
  scale_color_grey()

p
```

**06. For the previous data, create the histogram of the returns of the different stocks in different panels and save the result in a file named `'histograms.png'`.**

```{r}
p <- ggplot(df_stocks, aes(x = ret.adjusted.prices)) + 
  geom_histogram() + 
  facet_wrap(~ticker)

print(p)

# save in temp folder
ggsave(file.path(tempdir(), 'histograms.png'))
```

**07. Use function `tidyquant::tq_exchange` to find out all tickers from the NYSE exchange. Using `BatchGetSymbols::BatchGetSymbols`, download annual return data for all stocks in the exchange from 2010 until the current day. After that, create the mean/variance map by plotting the average annual return against its standard deviation. Tip: Use the parallel option at `BatchGetSymbols` to speed things up. You will find many outlier returns in the raw data. Make sure the plot is visible by limiting the x and y-axis (see functions `ggplot2::xlim` and `ggplot2::ylim`).**

```{r, cache=TRUE}
library(tidyquant)
library(BatchGetSymbols)
library(tidyverse)

df_nyse <- tq_exchange('NYSE')

my_tickers <- df_nyse$symbol

future::plan(future::multisession, 
             workers = floor(parallel::detectCores()/2))

df_prices <- BatchGetSymbols(tickers = my_tickers, first.date = '2010-01-01', 
                             last.date = Sys.Date(),
                             freq.data = 'yearly', 
                             do.parallel = TRUE)[[2]]

tab <- df_prices %>%
  na.omit() %>%
  group_by(ticker) %>%
  summarise(mean = mean(ret.adjusted.prices),
            sd = sd(ret.adjusted.prices))

p <- ggplot(tab, aes(x = sd, y = mean)) + 
  geom_point() + 
  xlim(0, 0.5) + ylim(-0.5, 0.5)

print(p)
```

**08. Head over to the [Kaggle data website](https://www.kaggle.com/datasets)^[https://www.kaggle.com/datasets] and choose a particular dataset for your analysis. It need not be related to economics or finance. Feel free to make a decision based on your own interests. After downloading the data, create a visual analysis of the data. Fell free to try out one or more plots of interest.  **

