---
title: "Analyzing Financial and Economic Data with R"
author: "Marcelo S. Perlin"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
  word_document: default
subtitle: "Chapter 03: Writing Research Scripts"
---

```{r setup, include=FALSE}
book.site <- 'https://www.msperlin.com/blog/publication/2017_book-pafdr-en/'
book.site.zip <- 'https://www.msperlin.com/blog/files/pafdr%20files/Code_Data_pafdR.zip'

knitr::opts_chunk$set(message = FALSE)
```


## Exercises

**01. Imagine a survey regarding your household budget over time. Data is stored as separate spreadsheets, one for each year, from 2009 to 2019. The objective of the research is to understand if it is possible to buy a home property in 5 years. Based on this, detail the elements in each stage of the study as a sequence, from importing the data to constructing the report. **

The possible stages of the study are:

1) Importing 01: Import all data regarding household income and budget, rate of return in safe (fixed income) investments and historical prices of home in desired locations;

2) Cleaning 01: Clean the data for outliers and bad (NA) cases;

3) Manipulation 01:Use the income data to find the personal ammount of cash saved each year.

4) Manipulation 02: Based on historical returns of fixed income investments and real state inflation, check how many years it takes to save the necessary ammount of cash to buy the real state. 

**Based on the previous exercise, create a folder structure on your computer to accommodate the research. Create dummy files with no content for each subdirectory. Note that the creation of the directories can be done with function `dir.create`.**

```{r}
library(fs)

# set temp dir for solving exercise
temp_dir <- path(tempdir(), 'Chapter 3 exercise')

# create folder
dir_create(temp_dir)

# create files
file_create(path(temp_dir, '01-import_data.R'))
file_create(path(temp_dir, '02-clean_data.R'))
file_create(path(temp_dir, '03-build_budget_table.R'))
file_create(path(temp_dir, '04-build_investment_realstate_table.R'))
file_create(path(temp_dir, '05-build_report.R'))

# create dirs
dir_create(path(temp_dir, 'data'))
dir_create(path(temp_dir, 'tabs'))
dir_create(path(temp_dir, 'figs'))

# fill with files
for (i_year in 2009:2019) {
  file_create(path(temp_dir, 'data', 
                   paste0('annual_budget_', i_year, '.xlsx')) )
}

file_create(path(temp_dir, 'data', 'fixed_income_index.csv'))
file_create(path(temp_dir, 'data', 'real_state_data.rds'))
file_create(path(temp_dir, 'figs', 'Income_and_HousePrices.png'))
file_create(path(temp_dir, 'tabs',  'Table_with_Results.xlsx'))

# Print a nice tree view of folder:
#   REQUIREMENT in Linux/Ubuntu: sudo apt install tree
#   Attention: it will NOT work in Windows

#cat(system(paste0('tree "', temp_dir, '"'), 
#           intern = TRUE), 
#    sep = '\n')
```

