---
title: "Analyzing Financial and Economic Data with R"
author: "Marcelo S. Perlin"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
  word_document: default
subtitle: "Chapter 04 - Importing and Exporting Data from Local Files"
---

```{r setup, include=FALSE}
book.site <- 'https://www.msperlin.com/blog/publication/2017_book-pafdr-en/'
book.site.zip <- 'https://www.msperlin.com/blog/files/pafdr%20files/Code_Data_pafdR.zip'

knitr::opts_chunk$set(message = FALSE)
```


## Exercises

**Create a dataframe with the following code:**

```{r, eval=FALSE}
my_N <- 10000
my_df <- data.frame(x = 1:my_N,
                    y = runif(my_N))
```

**Export the resulting dataframe to each of the following formats: csv, rds, xlsx e fst.**

**Which format took more computer space? Tip: You can use function `file.size` to check the resulting file sizes. **

```{r}
# csv
library(tidyverse)



do_tests <- function(my_N) {
  
  my_df <- tibble(x = 1:my_N,
                  y = runif(my_N))
  
  my_f <- tempfile(pattern = 'temp', fileext = '.csv')
  time.csv <- system.time({
    write_csv(my_df, my_f)
  })['elapsed']
  size.csv <- file.size(my_f)/1000000
  
  # rds
  my_f <- tempfile(pattern = 'temp', fileext = '.rds')
  time.rds <- system.time({
    write_rds(my_df, my_f)
  })['elapsed']
  
  size.rds <- file.size(my_f)/1000000
  
  # xlsx
  my_f <- tempfile(pattern = 'temp', fileext = '.xlsx')
  library(writexl) 
  time.xlsx <- system.time({
    write_xlsx(my_df, my_f)
  })['elapsed']
  
  size.xlsx <- file.size(my_f)/1000000
  
  # fst
  library(fst)
  my_f <- tempfile(pattern = 'temp', fileext = '.fst')
  time.fst <- system.time({
    write_fst(my_df, my_f)
  })['elapsed']
  
  size.fst <- file.size(my_f)/1000000
  
  print(c(size.csv, size.rds, size.xlsx, size.fst))
  
  print(c(time.csv, time.rds, time.xlsx, time.fst))
  
  tab <- tibble(Result = c('csv', 'rds', 'xlsx', 'fst'), 
                Size =  c(size.csv, size.rds, size.xlsx, size.fst),
                Time = c(time.csv, time.rds, time.xlsx, time.fst))
  
  return(tab)
}

my_N <- 10000

tab <- do_tests(my_N)
print(tab)

my_msg <- paste0('The format with largest disk space for N = ', my_N, ' is ', 
                 tab$Result[which.max(tab$Size)], '.')
cat(my_msg)

```

**02. Improve the previous code by measuring the execution time needed to record all data in the different formats. Which one had the minimum writing time? Tip: Use function `system.time` for measuring execution time. **

```{r}
my_msg <- paste0('The format with least execution time for N = ', my_N, ' is ', 
                 tab$Result[which.min(tab$Time)], '.')
cat(my_msg)
```


**03. Within the previous code, change the value of `my_N` to `1000000` and re-execute the code. Does this change modify the answers to the two previous questions? How?**

```{r}
my_N <- 1000000

tab <- do_tests(my_N)
print(tab)

my_msg <- paste0('The format with largest disk space for N = ', my_N, ' is ', 
                 tab$Result[which.max(tab$Size)], '.')

cat(my_msg)

my_msg <- paste0('The format with least execution time for N = ', my_N, ' is ', 
                 tab$Result[which.min(tab$Time)], '.')
cat(my_msg)
```


**04. Using functions `afedR::afedR_get_data_file` and `readr::read_csv`, read the contents of the book file `'SP500-Stocks_long.csv'`. How many columns and how many rows you find in this table?**

```{r, cache = TRUE}
library(tidyverse)

my_f <- afedR::afedR_get_data_file('SP500-Stocks_long.csv')

df_SP500 <- read_csv(my_f, 
                     col_types = cols())

my_msg <- paste0('There are ', nrow(df_SP500), ' rows and ', 
                 ncol(df_SP500), ' columns in file ', basename(my_f))
cat(my_msg)
```


**05. At link [https://eeecon.uibk.ac.at/~zeileis/grunfeld/Grunfeld.csv](https://eeecon.uibk.ac.at/~zeileis/grunfeld/Grunfeld.csv) you can find the CSV file for the _Grunfeld_ data. This is a particularly famous table due to its use as a benchmark data in econometric models. Using function `readr::read_csv`, read this file using the direct link in `read_csv`. How many rows do you find in the resulting `dataframe`?**

```{r}
library(tidyverse)

my_url <- 'https://eeecon.uibk.ac.at/~zeileis/grunfeld/Grunfeld.csv'

df_grunfeld <- read_csv(my_url, col_types = cols())

my_msg <- paste0('We find ', nrow(df_grunfeld), ' rows in the Grunfeld dataset.')
cat(my_msg)
```

**06. CHALLENGE - In the following link:**

[https://perso.telecom-paristech.fr/eagan/class/igr204/data/nat1900-2017.tsv](https://perso.telecom-paristech.fr/eagan/class/igr204/data/nat1900-2017.tsv)

**you can find data about all baby names in France from 1900 to 2017. But, the data comes with a caveat: all columns are separated by a tab symbol. After reading the manual for `readr::read_delim`, import the information from this file on your computer. How many rows does the file contain?**

```{r}
library(tidyverse)

my_url <- 'https://perso.telecom-paristech.fr/eagan/class/igr204/data/nat1900-2017.tsv'

df_names <- read_delim(my_url, delim = '\t', col_types = cols())

my_msg <- paste0('We find ', nrow(df_names), ' rows in the baby names dataset.')
cat(my_msg)
```

