---
title: "Analyzing Financial and Economic Data with R"
author: "Marcelo S. Perlin"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
  word_document: default
subtitle: Chapter 09 - Cleaning and Structuring Data
---

```{r setup, include=FALSE}
book.site <- 'https://www.msperlin.com/blog/publication/2017_book-pafdr-en/'
book.site.zip <- 'https://www.msperlin.com/blog/files/pafdr%20files/Code_Data_pafdR.zip'

knitr::opts_chunk$set(message = FALSE)
```

## Exercises

**01. Consider the following  `dataframe`:**

```{r}
library(tidyverse)

my_N <- 100

df <- bind_rows(tibble(ticker = rep('STOCK 1', my_N),
                       ref_date = Sys.Date() + 1:my_N,
                       price = 100 + cumsum(rnorm(my_N))),
                tibble(ticker = rep('STOCK 2', my_N),
                       ref_date = Sys.Date() + 1:my_N,
                       price = 100 + cumsum(rnorm(my_N))) )

print(df)
```

**The format is long or wide? Explain your answer.**

The data is in the long format. Only three columns can represent the dataset and, with the arrival of new stocks, rows should be appended. 

**02. Modify the format of the previous dataframe, from long to wide or vice-versa.**

```{r}
# convert from long to wide
df_wide <- spread(data = df, 
                  key = 'ticker',
                  value = 'price')

# print result
print(df_wide)
```


**03. Consider the following `list`:**

```{r}
my_l <- list(df1 = tibble(x = 1:100, y = runif(100)),
             df2 = tibble(x = 1:100, y = runif(100), v = runif(100)),
             df3 = tibble(x = 1:100, y = runif(100), z = runif(100)) )
```

**Aggregate all elements of `my_l` into a single dataframe. So, the question is, what happened to the data points in `df1` for columns `v` and `z`? **

```{r}
bind_df <- bind_rows(my_l)

print(bind_df)
```

After the aggregation, all columns that are not found in other `dataframes`, such as `v` and `z`, are set to `NA`.


**04. Use package `BatchGetSymbols` to download SP500 index data (`'^GSPC'`) from 1950-01-01 to today. What are the top 5 absolute returns (positive or negative) of the index? Create and present on screen a `dataframe` with the values and dates of these extreme returns.**


```{r}
library(BatchGetSymbols)

ticker <- '^GSPC'

df_SP500 <- BatchGetSymbols(tickers = '^GSPC', 
                            first.date = '1950-01-01',
                            last.date = Sys.Date())[[2]]

select_n <- 5
tab <- tibble(position = 1:select_n,
              top5_positive = sort(df_SP500$ret.adjusted.prices, 
                                   decreasing = TRUE)[1:select_n],
              top5_negative = sort(df_SP500$ret.adjusted.prices, 
                                   decreasing = FALSE)[1:select_n])

print(tab)
```


**05. Use the function `replace_outliers`, created in this chapter, for removing outliers from all columns of the SP500 data with `p = 0.025`. How many rows were lost in this process? **

```{r}
replace_outliers <- function(col_in, my_prob = 0.05) {
  # Replaces outliers from a vector
  #
  # INPUTS: col_in The vector
  #         my_prob Probability of quantiles (p and 1-p)
  #
  # OUTPUT: A vector
  
  # return if class is other than numeric
  if (!(class(col_in) %in% c('numeric', 'integer'))) return(col_in)
  
  my_outliers <- stats::quantile(x = col_in,
                                 probs = c(my_prob, 1-my_prob),
                                 na.rm = TRUE)
  
  idx <- (col_in <= my_outliers[1])|(col_in >= my_outliers[2])
  col_in[idx] <- NA
  
  return(col_in)
  
}

# remove outlivers from vectors
l_out <- map(df_SP500, replace_outliers, my_prob = 0.025)

df_SP500_nooutlier <- na.omit(as_tibble(l_out))

nrow_1 <- nrow(df_SP500)
nrow_2 <- nrow(df_SP500_nooutlier)

my_msg <- paste0('We lost ', nrow_1 - nrow_2, ' rows by removing the outliers.')
cat(my_msg)
```

**06. Use function `BatchGetSymbols::BatchGetSymbols` to download FTSE index prices (`'^FTSE'`) from `'2010-01-01'` to the present day. Next, build a dataset of index values in the annual frequency by looking at the latest available index value for each year. Tip: see function `dplyr::summarise_all` for a functional way of aggregating all columns.**

```{r}
library(BatchGetSymbols)

ticker <- '^FTSE'

df_FTSE_daily <- BatchGetSymbols(tickers = ticker, 
                            first.date = '2010-01-01',
                            last.date = Sys.Date())[[2]]

# change from daily to annual
df_FTSE_yearly <- df_FTSE_daily %>%
  mutate(ref_year = lubridate::year(ref.date)) %>%
  group_by(ref_year) %>%
  summarise(last_value = last(price.adjusted))

print(df_FTSE_yearly)
```

**07. Use the same daily data from previous exercise and build a new dataset in the monthly frequency.**

```{r}
# change from daily to monthly
df_FTSE_monthly <- df_FTSE_daily %>%
  mutate(ref_month = format(ref.date, '%Y-%m-01')) %>%
  group_by(ref_month) %>%
  summarise_all(last)

print(df_FTSE_monthly)
```


**08. CHALLENGE - For the previously downloaded FTSE daily data, check the dates and prices of the 20 biggest price drops. If an investor bought the index at the prices of the biggest drops and maintained it for 30 days, what would be his average nominal return per transaction?**

```{r}
library(tidyverse)

# buy at t, sell at t+30
trade_window <- 30 

# find largest drops
largest_drops <- df_FTSE_daily %>%
  arrange(ret.adjusted.prices) %>%
  slice(1:20)

# There are many ways to solve the exercise. 
# Here we will use a loop which is the simplest way to looking at the problem.
# You could also solve it with a functional approach of package purrrr

tab <- tibble()
for (i_date in seq_along(largest_drops$ref.date)) {
  
  my_date <- largest_drops$ref.date[i_date]
  # filter data to keep only datapoints in each horizon
  temp_df <- df_FTSE_daily %>%
    filter(ref.date >= my_date,
           ref.date <= my_date + trade_window)
  
  
  buy_price <- first(temp_df$price.adjusted)
  sell_price <- last(temp_df$price.adjusted)
  return <- sell_price/buy_price - 1
  
  tab <- bind_rows(tab, 
                   tibble(date = my_date, 
                          buy_price = buy_price, 
                          sell_price = sell_price, 
                          return = return))
}

print(tab)

# solution
avg_ret <- scales::percent(mean(tab$return))

my_msg <- paste0('The average return per transaction is ', avg_ret, '.')
cat(my_msg)
```

