```{r data generation, echo = FALSE, results = "hide"}
file_football <- afedR_get_data_file('Brazil_footbal_games.csv')
df <- readr::read_csv(file_football, col_types = readr::cols(),
                      locale = locale(encoding = 'Latin1'))

tab <- df  %>% 
  group_by(Adversario) %>%
  summarise(nwin = sum(GolsBR>GolsAdv)) %>%
  arrange(-nwin, Adversario)

most.losses <- tab$Adversario[1]

temp.df <- df[df$Adversario == most.losses, ] %>%
  arrange(Data) %>%
  slice(n())

sol.q <- paste0(temp.df$GolsBR, ' - ', 
                temp.df$GolsAdv)

temp <- paste0(df$GolsBR, ' - ', 
               df$GolsAdv)
possible.anwers <- temp[temp != sol.q]
  
rnd.vec <- c(sol.q, sample(possible.anwers, 4) )
my.answers <- rnd.vec
```


Question
========

For the same data found within file `r file_football`, what is the score (BR X ADV) of the last game against the team with the most number of defeats to the Brazilian team? Once again, in case of draw, use the team by alphabetical order.

```{r questionlist, echo = FALSE, results = "asis"}
answerlist(my.answers, markup = "markdown")
```


Meta-information
================
extype: schoice
exsolution: `r mchoice2string(c(TRUE, FALSE, FALSE, FALSE, FALSE), single = TRUE)`
exname: t statistic
exshuffle: TRUE
