```{r data generation, echo = FALSE, results = "hide"}
require(tidyverse)

my.sp500.f <- afedR::afedR_get_data_file('SP500_comp_YEARLY_long_2014-10-03_2019-10-02.rds')
df.sp500 <- readr::read_rds(my.sp500.f)

tab <- df.sp500 %>%
  dplyr::group_by(ticker) %>%
  dplyr::summarise(mean.ret = mean(ret.adjusted.prices, na.rm = TRUE),
            sd.ret = sd(ret.adjusted.prices, na.rm = TRUE))

sol.q.1 <- tab$ticker[which.min(tab$sd.ret)]
sol.q.2 <- tab$ticker[which.max(tab$sd.ret)]

idx <- !(tab$ticker %in% c(sol.q.1, sol.q.2))
possible.anwers <- tab$ticker[idx]

sol.q <- paste0(sol.q.1, ' e ', sol.q.2)

#rnd.vec <- gen.rnd.vec()
#my.answers <- format(rnd.vec*sol.q , digits = 4)

temp <- paste0(sample(possible.anwers, 4), ' e ', sample(possible.anwers, 4))

my.answers <- c(sol.q, temp)
```

Question
========

Open RStudio and load the data from file `r my.sp500.f` using `afedR::afedR_get_data_file`. This file contains stock data for companies in the SP500 index in the yearly frequency. Using the programming tools of R, create a new table with three columns:

- stock's ticker
- average yearly return;
- standard deviation of yearly returns.

Which stocks have the lowest and highest standard deviation, respectively?

Note: You'll have to deal with NA values and remove them within the originial data. 


```{r questionlist, echo = FALSE, results = "asis"}
answerlist(my.answers, markup = "markdown")
```


Meta-information
================
extype: schoice
exsolution: `r mchoice2string(c(TRUE, FALSE, FALSE, FALSE, FALSE), single = TRUE)`
exname: t statistic
exshuffle: TRUE
