```{r data generation, echo = FALSE, results = "hide"}
require(readr)
require(tidyverse)

my_file <- afedR_get_data_file('FTSE.csv')
df_index <- readr::read_csv(my_file, col_types = readr::cols())

tab <- df_index  %>%
  mutate(weekday = weekdays(ref.date)) %>%
  group_by(weekday) %>%
  summarise(mean.vol = mean(volume))

df_index <- df_index %>%
   mutate(weekday = weekdays(ref.date))

min.vol <- tab$weekday[which.min(tab$mean.vol)]
max.vol <- tab$weekday[which.max(tab$mean.vol)]

my.ttest <- t.test(df_index$volume[df_index$weekday == min.vol],
                   df_index$volume[df_index$weekday == max.vol])

sol.q <- my.ttest$p.value

rnd.vec <- afedR::afedR_gen_rnd_vec()
my.answers <- format(rnd.vec*sol.q, digits = 4)
```

Question
========

Using the same data from previous exercise, file `r basename(my_file)`, test if the different of means from the traded volume of the day of the week with the lowest average value agasint the week day with the highest average value of trading volume. For that, use a simple T-test (function `t.test`).

What is the p-value of the test?

```{r questionlist, echo = FALSE, results = "asis"}
answerlist(my.answers, markup = "markdown")
```

Meta-information
================
extype: schoice
exsolution: `r mchoice2string(c(TRUE, FALSE, FALSE, FALSE, FALSE), single = TRUE)`
exname: t statistic
exshuffle: TRUE
