```{r data generation, echo = FALSE, results = "hide"}
file_football <- afedR_get_data_file('Brazil_footbal_games.csv')
df <- readr::read_csv(file_football, col_types = readr::cols(),
                      locale = locale(encoding = 'Latin1'))

tab <- df  %>% 
  group_by(Adversario) %>%
  summarise(nwin = sum(GolsBR<GolsAdv)) %>%
  arrange(-nwin, Adversario)

tab  
sol.q <- tab$Adversario[1]

rnd.vec <- c(sol.q, sample(tab$Adversario[tab$Adversario != sol.q], 4) )
my.answers <- rnd.vec
```


Question
========

Load the data from file `r basename(file_football)` using `afedR::afedR_get_data_file. This file contains data taken from Google about games of the national football team. 

Using loops or the `dplyr` package, check which teams accumulated the highest number of victories over the national team? In case of a draw, use the alphabetical order of team names for the solution. 


```{r questionlist, echo = FALSE, results = "asis"}
answerlist(my.answers, markup = "markdown")
```

Meta-information
================
extype: schoice
exsolution: `r mchoice2string(c(TRUE, FALSE, FALSE, FALSE, FALSE), single = TRUE)`
exname: soccer most wins
exshuffle: TRUE
