```{r data generation, echo = FALSE, results = "hide"}
require(readr)
require(tidyverse)

my_file <- afedR_get_data_file('FTSE.csv')
df_index <- readr::read_csv(my_file, col_types = readr::cols())

tab <- df_index %>%
  mutate(weekday = weekdays(ref.date)) %>%
  group_by(weekday) %>%
  summarise(mean.volume = mean(volume))

sol.q <- tab$weekday[which.min(tab$mean.volume)]

rnd.vec <- c(sol.q, sample(tab$weekday[tab$weekday != sol.q],4))
my.answers <- rnd.vec
```

Question
========

Import the data available at file `r basename(my_file)` using function `afedR::afedR_get_data_file`.  There you will find dates, prices and traded volume for the FTSE index. Now, calculate the average trading volume for each weekday (monday, tuesday, ..). What is the week day with the lowest traded volume?

```{r questionlist, echo = FALSE, results = "asis"}
answerlist(my.answers, markup = "markdown")
```


Meta-information
================
extype: schoice
exsolution: `r mchoice2string(c(TRUE, FALSE, FALSE, FALSE, FALSE), single = TRUE)`
exname: volume and day of week
exshuffle: TRUE
