```{r data generation, echo = FALSE, results = "hide"}
file_football <- afedR::afedR_get_data_file('Brazil_footbal_games.csv')
df.football <- readr::read_csv(file_football, col_types = readr::cols(),
                      locale = readr::locale(encoding = 'Latin1'))

my.file.ibov <- afedR::afedR_get_data_file('Ibov_long_2010-01-01_2018-09-12.rds')
df.ibov <- readr::read_rds(my.file.ibov)

games.dates <- as.Date(df.football$Data)
ibov.dates <- as.Date(df.ibov$ref.date)
diff.goals <- df.football$GolsBR - df.football$GolsAdv

find.idx.date <- function(my.date, ibov.dates, type){
  
  dist <- ibov.dates - my.date  
  
  if (type == 'bef'){
    dist[dist>=0] <- NA
    idx <- which.min(abs(dist))
    
  } 
  
  if (type == 'aft'){
    dist[dist<=0] <- NA
    idx <- which.min(abs(dist))
    
  } 
  
  return(idx)
  
}

idx.aft <-sapply(games.dates, FUN = find.idx.date, ibov.dates=ibov.dates, type='aft')
idx.bef <-sapply(games.dates, FUN = find.idx.date, ibov.dates=ibov.dates, type='bef')

after.date <- ibov.dates[idx.aft]
bef.date <- ibov.dates[idx.bef]
ret.after.date <- df.ibov$ret.adjusted.prices[idx.aft]
ret.before.date <- df.ibov$ret.adjusted.prices[idx.bef]


df.answer <<- data.frame(games.dates = games.dates, 
                        diff.goals = diff.goals,
                        after.date = after.date,
                        ret.after.date = ret.after.date,
                        bef.date=bef.date,
                        ret.before.date = ret.before.date)
                        
sol.q <- sum(df.answer$ret.after.date>0, na.rm = T)

rnd.vec <- afedR::afedR_gen_rnd_vec()
my.answers <- format(floor(rnd.vec*sol.q), digits = 4)

```

Question
========

Using previous tables, files `r basename(file_football)`) and `r basename(my.file.ibov)`, create a table where the lines are related to each game and the columns are:

- date of game; 
- difference of goals (BR minus Adv); 
- the posterior date where the market was open; 
- return of the Ibovespa index in the posterior date; 
- the previous date where the market was open; 
- return of the Ibovespa index in the previous date.

Question: what is the quantity of positive returns in the fourth column of the resulting table?

Tip: The simplest way to solve the problem is using loops. Iterate each game/row of the main table and do the required calculations.

```{r questionlist, echo = FALSE, results = "asis"}
answerlist(my.answers, markup = "markdown")
```


Meta-information
================
extype: schoice
exsolution: `r mchoice2string(c(TRUE, FALSE, FALSE, FALSE, FALSE), single = TRUE)`
exname: t statistic
exshuffle: TRUE

